---
title: "Stawberries 3"
author: MA615
date: 2025 October 17
format: 
    html:
       embed-resources: false
editor: source 
---

# A comprehensive project

Data discovery, acquisition, cleaning, organization, and exploration

# Preparing data for analysis

Research, discovery, acquisition, clean, organize, explore, report

## Research

Main data source: [USDA Quickstats](https://quickstats.nass.usda.gov/)

[Census of Agriculture](https://www.nass.usda.gov/AgCensus/)

[Survey of Agriculture](https://www.nass.usda.gov/Surveys/index.php)

North American Strawberry Growers Association: <https://www.nasga.org/>)

Organic Farmers Association: <https://organicfarmersassociation.org/>

[US Strawberry Market Annual Report, June 20, 2024](https://www.freshproduce.com/siteassets/files/reports/global-trade/2024/strawberries_annual_market_report_2024.pdf)

<https://www.freshproduce.com/siteassets/files/reports/global-trade/2024/strawberries_annual_market_report_2024.pdf>

<https://www.agmrc.org/commodities-products/fruits/strawberries>

["An introduction to data cleaning with R" by Edwin de Jonge and Mark van der Loo](https://cran.r-project.org/doc/contrib/de_Jonge+van_der_Loo-Introduction_to_data_cleaning_with_R.pdf)

["Problems, Methods, and Challenges in Comprehensive Data Cleansing" by Heiko Müller and Johann-Christoph Freytag](https://www.researchgate.net/profile/Heiko-Mueller/publication/228929938_Problems_methods_and_challenges_in_comprehensive_data_cleansing/links/09e415101b58541e2c000000/Problems-methods-and-challenges-in-comprehensive-data-cleansing.pdf)

## Strawberries

**Questions**

-   Where they are grown? By whom?

-   Are they really loaded with carcinogenic poisons?

-   Are they really good for your health? Bad for your health?

-   Are organic strawberries carriers of deadly diseases?

-   When I go to the market should I buy conventional or organic strawberries?

-   Do Strawberry farmers make money?

-   How do the strawberries I buy get to my market?

## The data

The data set for this assignment has been selected from:

## read and explore the data

Set-up

```{r}
#| label: load libraries and set options
#| warning: false
#| message: false

library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)



```

Read the data and take a first look

```{r}
#| label: read data - glimpse 

strawberry <- read_csv("strawberry_10Oct25.csv", col_names = TRUE)

# glimpse(strawberry)
##Oct25
```



All I can see from the glimpse is I have date, location, values and coefficients of variation.

Examine the data. How is it organized?

```{r}
#| label: explore organization 1 

## Is every line associated with a state?

state_all <- strawberry |> distinct(State)

state_all1 <- strawberry |> group_by(State) |> count()

years <- strawberry |> distinct(Year)


## every row is associated with a state

if(sum(state_all1$n) == dim(strawberry)[1]){print("Yes every row in the data is associated with a state.")}

## rm(state_all, state_all1)

```

## remove columns with a single value in all rows

```{r}
#|label: function def - drop 1-item columns

drop_one_value_col <- function(df){   ## takes whole dataframe
drop <- NULL  

## test each column for a single value
for(i in 1:dim(df)[2]){     
if((df |> distinct(df[,i]) |> count()) == 1){
drop = c(drop, i)
} }

## report the result -- names of columns dropped
## consider using the column content for labels 
## or headers 

if(is.null(drop)){return("none")}else{

   print("Columns dropped:")
   print(colnames(df)[drop])
   strawberry <- df[, -1*drop]
   }
}


## use the function

strawberry <- drop_one_value_col(strawberry)

drop_one_value_col(strawberry)

```

To get better look at the data, look at California.

```{r}
#| label: explore California only

calif <- strawberry |>  filter(State=="CALIFORNIA")

## look at the unique values in the "Program" column

## in the consol
## unique(calif$Program)

## and look at the data selection widget on 
##      https://quickstats.nass.usda.gov

## You can see that CENSUS AND SURVEY are the two sources
## of data. (Why? What's the differences?).  So, let's see
## they differ.

calif_census <- calif |> filter(Program=="CENSUS")

calif_survey  <- calif |>  filter(Program=="SURVEY")

### 


##calif_survey <- strawberry |> select(Year, Period, `Data Item`, Value)


```

Explore California to understand the census and survey

```{r}
#| label: explore Calif census and survey 

## no assignment -- just exploring

drop_one_value_col(calif_census)

drop_one_value_col(calif_survey)



```

Conclusions from California data exploration.


Clean up the variables created while exploring

```{r}
#| label: cleanup after CA

rm(calif, calif_census, calif_survey, 
   state_all, state_all1)

```




## Now return to the entire data set.

take the lessons learned by examinging the California data

Before we split the CENSUS rows from the SURVEY rows,
here is a "low hanging fruit" opportunity ...

```{r}
#| label: test for a potential one-value col

## look at col = `Data Item`
## looks like every row starts with "STRAWBERRIES" 
## TEST IT 

test1 <- str_starts(strawberry$`Data Item`, 
                    "STRAWBERRIES - " )

sum(test1) == dim(strawberry)[1]

test2 <- str_starts(strawberry$`Data Item`, "STRAWBERRIES" )
sum(test2) == dim(strawberry)[1]

## ok sould you drop the 
## leading "STRAWBERRIES" from each row?
## that depends on if you understand all the patterns


## try two examples 
t1 <- which(test1 == TRUE)[2]
strawberry$`Data Item`[t1]

t2 <- which(test1 == FALSE)[2]
strawberry$`Data Item`[t2]

##
## Use str_detect to build a regex pattern 
## that matches both instances

str_detect(strawberry$`Data Item`[t1], "^ST.*IES\ -\ ")

str_detect(strawberry$`Data Item`[t1], "^ST.*IES(,\ )")

## the expressions above may work as an 'or'

str_detect(strawberry$`Data Item`[t2], 
           "^ST.*IES((\ -\ )|(,\ ))")

## test it

test3 <- str_starts(strawberry$`Data Item`, 
                    "^ST.*IES((\ -\ )|(,\ ))")

sum(test3) == dim(strawberry)[1]

## OK use the regex to cleanup the 
## `Date Item` column

strawberry$`Data Item` <- 
  str_replace(strawberry$`Data Item`, 
              "^ST.*IES((\ -\ )|(,\ ))", "")

## and clean up the workspace 
rm(t1, t2, test1, test2, test3)

```

OK -- 
Split the census dat from the survey data. drop single value columns

```{r}
#| label: survey, census spit

strawb_survey <- strawberry |> filter(Program=="SURVEY")

strawb_census <- strawberry |> filter(Program=="CENSUS")

strawb_survey <- drop_one_value_col(strawb_survey)

strawb_census <- drop_one_value_col(strawb_census)


```





```{r}
#| label: census data item analysis

census_data_items <- strawb_census |> distinct(`Data Item`)

```


Use census data items to plan how to use and show the information in the ag census data.



## separate composite columns

### Now examine strawb_survey

```{r}
#|label: examine strawb_survey

surv_data_item <- strawb_survey |> distinct(`Data Item`)

## observe the data 
## we want so split the data category 
## from the data

t1 <- str_detect(surv_data_item$`Data Item`, "\ -\ ")

strawb_survey$`Data Item` <- str_replace(strawb_survey$`Data Item`, ", " ,"\ -\ ")


strawb_survey <- strawb_survey |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = " - ",
                         names = c("category",
                                 "data1",
                                 "data2"),
                         too_many = "error",
                         too_few = "align_start"
                       )


```


You can see that the chemical data is not in this data set. 
What do you do?

#### usda nass system  https://quickstats.nass.usda.gov/

or

#### R Package rnassqs 

